Negative Selection in the 20th Century

### 2020-06-12 why-negative-selection.Rmd
Updates to graphs in response to Abdel's comments

drake::loadd(rgs)
drake::loadd(score_names)

@@

theme_set(theme_minimal())
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "orange2")

@@

my_fill_scale <- function (n, aesthetics = "fill", ...) {
  if (n == 2) {
    scale_colour_manual(values = c("steelblue4", "indianred2"), 
          aesthetics = aesthetics, ...)
    # scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

pretty_names <- function (names) {
  names <- gsub("_", " ", names)
  names %<>% 
        strsplit(" ") %>% 
        map_chr(~paste0(
          toupper(substring(.x, 1, 1)), 
          substring(.x, 2),
          collapse = " "
        ))

  acronyms <- c("WHR", "BMI", "DPW", "CPD", "SI", "AI", "WC")
  for (acronym in acronyms) {
    regex <- paste0("\\b", acronym, "\\b")
    names <- gsub(regex, toupper(acronym), names, ignore.case = TRUE)
  }

  return(names)
}

@@

drake::loadd(famhist)
drake::loadd(ghs_weights)
famhist <- left_join(famhist, ghs_weights, by = "f.eid")
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

@@

```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, score, group = score_name)) + 
      geom_line(colour = "steelblue4") +
      facet_wrap(~score_name) +
      my_hline +
      theme(
        strip.text  = element_text(size = 7), 
        axis.text.x = element_text(size = 7)
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")
```

@@

Estimates are broadly consistent across generations. However, effect sizes are 
much larger for *siblings* than for *children* 
regressions. Among consistently-signed estimates, the effect size for
siblings is `r 100*median_prop_gens` per cent at the median.

@@

```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}

@@

res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

drake::loadd(res_edu)
n_regs <- as.double(nrow(res_edu))
res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
      ) %>% 
      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
        fill = `Age left FTE`)) +
        my_vline +
        pval_geom_point() + 
        pval_scale() +
        my_fill_scale(n = 3, direction = -1) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        )
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = Income, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point(alpha = 0.8) + 
        my_fill_scale(n = 5, direction = -1) +
        pval_scale() +
        my_labs() +
        my_theme()
```

@@

res_age_flb_cross %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1)),
        term       = gsub("age_flb_cat(.*):.*", "\\1", term)
      ) %>% 
      ggplot(aes(estimate, score_name, fill = term, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 3, direction = -1, 
          guide = guide_legend(title = "Age at FLB")) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      ggplot(aes(y = reorder_within(score_name, estimate, sex, fun = order_abs(1)), x = estimate, fill = `N partners`, 
        colour = p.value < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        my_theme(legend.position = "bottom") +
        my_labs(x = "", y = "") + 
        scale_y_reordered()

@@

n_regs <- nrow(res_with_partner)

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", "Without partner")
      ) %>% 
      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
```

@@

```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
famhist %>% 
      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
      mutate(
        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
        "N Children"  = santoku::chop(n_children, brk_left(0:4, FALSE), 
                          labels = c(0:3, "4+")),
        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
        ) %>% 
      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
        facet_wrap(vars(Sex)) +
        scale_fill_viridis_d(option = "A") +
        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
        theme(legend.position = "bottom")

>>>>

drake::loadd(rgs)
drake::loadd(score_names)
drake::loadd(famhist)

@@

theme_set(theme_minimal())
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")

@@

my_fill_scale <- function (n, aesthetics = "fill", ...) {
  if (n == 2) {
    scale_colour_manual(values = c("steelblue4", "#D63C3C"), 
          aesthetics = aesthetics, ...)
    # scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

pretty_names <- function (names) {
  pretty <- c(
    ADHD_2017               = "ADHD",
    age_at_menarche         = "Age at menarche",
    age_at_menopauze        = "Age at menopause",
    agreeableness           = "Agreeableness",
    ai_substance_use        = "Age at smoking initiation",
    alcohol_schumann        = "Alcohol use",
    alzheimer               = "Alzheimer",
    autism_2017             = "Autism",
    bipolar                 = "Bipolar",
    bmi_combined            = "BMI",
    body_fat                = "Body Fat",
    caffeine                = "Caffeine",
    cannabis                = "Cannabis (ever vs. never)",
    cognitive_ability       = "Cognitive Ability",
    conscientiousness       = "Conscientiousness",
    coronary_artery_disease = "Coronary Artery Disease",
    cpd_substance_use       = "Cigarettes per day",
    diagram_T2D             = "Type 2 Diabetes",
    dpw_substance_use       = "Drinks per week",
    EA2_noUKB               = "Educ. attainment 2 (no UK)",
    EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
    eating_disorder         = "Eating disorder",
    extraversion            = "Extraversion",
    height_combined         = "Height",
    hip_combined            = "Hip circumference",
    MDD_PGC2_noUKB          = "Major Depressive Disorder",
    neuroticism             = "Neuroticism",
    openness                = "Openness",
    sc_substance_use        = "Smoking cessation",
    SCZ2                    = "Schizophrenia",
    si_substance_use        = "Smoking initiation",
    wc_combined             = "Waist circumference",
    whr_combined            = "Waist-hip ratio"
  )

  pretty[names]
}

n_in_regs <- function (var, dv = "n_children") {
  # all PGS have the same NA pattern, so we just use one
  tbl <- table(famhist[
                  ! is.na(famhist[dv]) & ! is.na(famhist$whr_combined), 
                  var
                ])
  c(tbl)
}

add_n <- function (var, famhist_var = var, dv = "n_children") {
  var <- as.factor(var)
  n <- n_in_regs(famhist_var, dv)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[levels(var)], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

@@

drake::loadd(famhist)
drake::loadd(ghs_weights)
famhist <- left_join(famhist, ghs_weights, by = "f.eid")
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          seq(1940, 1970, 5), 
          labels    = seq(40, 65, 5),
          extend    = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
      mutate(
        score_lo = score$ymin,
        score_hi = score$ymax,
        score    = score$y
      )

@@

```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals."}

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        colour = "steelblue4", fatten = 2) +
      facet_wrap(~score_name) +
      my_hline +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7)
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

Estimates are broadly consistent across generations. However, effect sizes are 
much larger for *siblings* than for *children* 
regressions. Among consistently-signed estimates, at the median, the proportion between
effect sizes for siblings and for children is `r 100 * median_prop_gens` per cent.

@@

```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex. Lines show significant differences"}

@@

res_sex_wide <- res_sex %>% 
      select(term, estimate, sex, diff.p.value) %>% 
      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
      mutate(
        min_est = pmin(Female, Male),
        max_est = pmax(Female, Male),
        term = pretty_names(term),
        term = fct_reorder(term, Female)
      )

res_sex %>% 
      mutate(
        term = pretty_names(term),
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        geom_linerange(data = res_sex_wide, 
          aes(
            y      = term,
            x      = NULL,
            fill   = NULL,
            xmin   = min_est, 
            xmax   = max_est, 
            colour = diff.p.value < 0.05/{{n_regs}}
          ), 
        group = "") +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

drake::loadd(res_edu)
n_regs <- as.double(nrow(res_edu))

res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
      ) %>% 
      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
        fill = `Age left FTE`)) +
        my_vline +
        pval_geom_point() + 
        pval_scale() +
        my_fill_scale(n = 3, direction = -1) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        ),
        Income = add_n(Income, "income_cat")
      ) %>% 
      ggplot(aes(y = term, x = estimate, fill = Income, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point(alpha = 0.8) + 
        my_fill_scale(n = 5, direction = -1) +
        pval_scale() +
        my_labs() +
        my_theme()
```

@@

res_age_flb_cross %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1)),
        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
        term       = add_n(term, "age_flb_cat")
      ) %>% 
      ggplot(aes(estimate, score_name, fill = term, 
        colour = p.value < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 3, direction = -1, 
          guide = guide_legend(title = "Age at first live birth")) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p.value < 0.05/{{n_regs}}
      )) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        my_theme(legend.position = "bottom") +
        my_labs(x = "", y = "") + 
        scale_y_reordered()

@@

n_regs <- nrow(res_with_partner)
hhtype_n <- n_in_regs("with_partner")

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        with_p    = as.character(grepl("TRUE", term)),
        Household = sprintf("%s (N = %s)", Household, 
                      prettyNum(hhtype_n[with_p], big.mark = ","))
      ) %>% 
      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
```

@@

```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
famhist %>% 
      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
      mutate(
        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
        "N Children"  = santoku::chop(n_children, 0:4, labels = c(0:3, "4+")),
        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
        ) %>% 
      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
        facet_wrap(vars(Sex)) +
        scale_fill_viridis_d(option = "A") +
        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
        theme(legend.position = "bottom")
